Ecosystem size and complexity dictate riverine biodiversity

### 2021-01-15 manuscript.Rmd
Update manuscript

---
title: "Ecosystem size and complexity dictate riverine biodiversity"
author:
- Akira Terui^[Department of Biology, University of North Carolina at Greensboro]
- Seoghyun Kim\footnotemark[1]
- Christine L. Dolph^[Department of Ecology, Evolution and Behavior, University of Minnesota]
- Taku Kadoya^[National Institute for Environmental Studies]
- Yusuke Miyazaki^[Shiraume Gakuen University]
output:
  word_document: default
  html_notebook: default
abstract:
  The species-area relationship (the SAR), a tendency in which species richness increases with increasing area, is one of the most robust empirical generalizations in ecology and has provided the foundation for area-based conservation strategies. However, many ecosystems possess complex spatial structure that cannot be repserented by area (the scale-invariant complexity) and its role in scaling biodiversity is largely unknown. Here, we hypthesized that ecosystem size and complexity jointly regulate biodiversity in rivers, a prime example of fractal branching networks where the recurrent merging of diverse streams generates a pronounced self-similarity in complex branching patterns. Ecosystem size should dictate metacommunity size and habitat diversity that regulate watershed-scale species richness. Meanwhile, ecosystem complexity, which we defined as the probability of branching per unit river distance, should increase watershed-scale species richness by increasing habitat heterogeneity. We developed theoretical models simulating metacommunity dynamics in branching river networks and yielded results consistent with our hypothesis. Ecosystem size and complexity both increased watershed-scale species richness ($\gamma$ diversity), and the importance of spatial turnover in species composition ($\beta$ diversity) was apparent when simulated species travel short distances. Our cross-watershed comparisons across two biogeographic regions (Hokkaido, Japan and Midwest, US) provided empirical supports for the theoretical predictions. As predicted, watershed-scale species richness increased with ecosystem size (watershed area) and complexity (branching probability) through enhanced spatial turnover in species composition over the branching network. There is now a clear need to consider the dimension of ecosystem complexity in biodiversity conservation. The dual dimensional approach - size and complexity - should broaden viable options for spatial planning of protected areas or restoration sites, thereby helping reconcile the conflicts between biodiversity conservation and societal demands.

@@

First, we theoretically analyzed the influences of ecosystem size and branching complexity on $\gamma$ diversity. We constructed a general simulation model of metacommunity dynamics [@thompsonProcessbasedMetacommunityFramework2020], in which 50 competing species disperse and spatially interact in a random branching network. Simulated species with different niche in abiotic conditions (optimum and width) compete for resources with varied strengths. Here, we depicted branching ecosystems as a spatial network of connected habitat patches (**Figure 1**) where mean environmental conditions at each habitat patch are determined by two factors: (1) headwater environments (the most upstream habitat patch) and (2) local environmental noise. Environmental values at headwaters are drawn randomly from a normal distribution and propagate downstream with local environmental noise (i.e., a spatial autoregressive process with white noise). These environmental values recurrently 'mix' at confluences considering the relative size of joining tributaries (see Methods). Therefore, our simulation resembles natural processes of how branching river networks create diverse habitats in a metacommunity. We used `r n_scenario` sets of parameter combinations with varied ecological (dispersal and competition strength) and landscape scenarios (the environmental variation at headwaters and the degree of local environmental noise). Under each scenario, we simulated metacommunity dynamics in `r max(dat_sim$n_rep)` branching networks that have different ecosystem size (the number of habitat patches: `r min(dat_sim$n_patch)` to `r max(dat_sim$n_patch)`) and complexity (branching probability: `r round(min(dat_sim$p_branch), 2)` to `r round(max(dat_sim$p_branch), 2)`) to examine their effects on species diversity.

@@

$$
\begin{align}
\mu_{z,k} &= \omega(\rho \mu_{z,k+1}^R + \epsilon_{k}^R) + (1-\omega)(\rho \mu_{z,k+1}^L + \epsilon_{k}^L) \\
\omega &= \frac{A_{k+1}^R}{A_{k+1}^R + A_{k+1}^L}
\end{align}
$$
where $R$ and $L$ denote "right" and "left" branches, respectively, and $A_k$ is the number of upstream habitat patches at patch $k$ (akin to upstream watershed area). With this expression, larger tributaties have a greater influence on the downstream environment, as observed in natural river networks[@bendaNetworkDynamicsHypothesis2004].

@@

We used fixed values for the following parameters: maximum reproductive number ($r_{0,i} =$ 4), niche cost ($\nu =$ `r unique(dat_sim$niche_cost)`), the degree of temporal fluctuation in abiotic environments ($\sigma_z =$ `r unique(dat_sim$sd_env)`), and the degree of spatial autocorrelation in temporal environmental dynamics ($\phi =$ `r unique(dat_sim$phi)`).

@@

We selected data based on the following criteria. First, we selected sampling sites with reliable coordinates. Sampling sites were treated as the same site if coordinates are identical at 3 decimal degrees. Second, for sites with multiple site visits (i.e., temporal replicates), we used the latest data at each sampling site to minimize variation in sampling efforts among sites. Lastly, we selected metacommunities (i.e., watersheds) that contain more than 10 sampling sites to reduce statistical uncertainties in estimating $\gamma$ diversity. We considered watersheds as separated from others if the river network flows into one of the following: the ocean, a large lake ($\ge$ 10 km² in the areal area) or a large river that may represent lentic habitats ($\ge$ 5000 km² in watershed area).

@@

We collected data for watershed characteristics (watershed area and branching probability), climatic variables (annual mean air temperature and cumulative precipitation), and land use (% forest). We delineated watershed polygons and riverlines using Hydrologically Adjusted Elevations of MERIT Hydro[@yamazakiMERITHydroHighResolution2019] with ArcMap 10.7 and SAGA 6.3.0 combined with an R package 'RSAGA'[@brenningRSAGASAGAGeoprocessing2018]. Watershed area was estimated based on the delineated watershed polygons using functions in an R package 'sf'[@pebesmaSimpleFeaturesStandardized2018]. Branching probability was estimated by fitting exponential distributions to histograms of branch length (unit: kilometer, river segment between successive confluences or a confluence and the outlet/upstream terminal) as $branch~length \sim Exp(\lambda)$. A derived statistical quantity $1 − e^{−\lambda}$ corresponds to the theoretical branching probability $P_b$. For each watershed, we estimated the spatial averages of air temperature and cumulative precipitation at a 1-km resolution, which were extracted from WorldClim 1.4[@hijmansVeryHighResolution2005] via R package 'raster'. We calculated % forest for each watershed using land use data in 2015 from Copernicus Global Land Service (100-m resolution)[@marcelbuchhornCopernicusGlobalLand2020].

@@

>>>>

---
title: "Ecosystem size and complexity dictate riverine biodiversity"
author:
- Akira Terui^[Department of Biology, University of North Carolina at Greensboro]
- Seoghyun Kim\footnotemark[1]
- Christine L. Dolph^[Department of Ecology, Evolution and Behavior, University of Minnesota]
- Taku Kadoya^[National Institute for Environmental Studies]
- Yusuke Miyazaki^[Shiraume Gakuen University]
output:
  html_notebook: default
  word_document: default
abstract:
  The species-area relationship (the SAR), a tendency in which species richness increases with increasing area, is one of the most robust empirical generalizations in ecology and has provided the foundation for area-based conservation strategies. However, many ecosystems possess complex spatial structure that cannot be repserented by area (the scale-invariant complexity) and its role in scaling biodiversity is largely unknown. Here, we hypthesized that ecosystem size and complexity jointly regulate biodiversity in rivers, a prime example of fractal branching networks where the recurrent merging of diverse streams generates a pronounced self-similarity in complex branching patterns. Ecosystem size should dictate metacommunity size and habitat diversity that regulate watershed-scale species richness. Meanwhile, ecosystem complexity, which we defined as the probability of branching per unit river distance, should increase watershed-scale species richness by increasing habitat heterogeneity. We developed theoretical models simulating metacommunity dynamics in branching river networks and yielded results consistent with our hypothesis. Ecosystem size and complexity both increased watershed-scale species richness ($\gamma$ diversity), and the importance of spatial turnover in species composition ($\beta$ diversity) was apparent when simulated species travel short distances. Our cross-watershed comparisons across two biogeographic regions (Hokkaido, Japan and Midwest, US) provided empirical supports for the theoretical predictions. As predicted, watershed-scale species richness increased with ecosystem size (watershed area) and complexity (branching probability) through enhanced spatial turnover in species composition over the branching network. There is now a clear need to consider the dimension of ecosystem complexity in biodiversity conservation. The dual dimensional approach - size and complexity - should broaden viable options for spatial planning of protected areas or restoration sites, thereby helping reconcile the conflicts between biodiversity conservation and societal demands.

@@

First, we theoretically analyzed the influences of ecosystem size and branching complexity on $\gamma$ diversity. We constructed a general simulation model of metacommunity dynamics [@thompsonProcessbasedMetacommunityFramework2020], in which 50 competing species disperse and spatially interact in a random branching network. Simulated species with different niches in abiotic conditions (optimum and width) compete for resources with varied strengths. Here, we depicted branching ecosystems as a spatial network of connected habitat patches (**Figure 1**) where mean environmental conditions at each habitat patch are determined by two factors: (1) headwater environments (the most upstream habitat patch) and (2) local environmental noise. Environmental values at headwaters are drawn randomly from a normal distribution and propagate downstream with local environmental noise (i.e., a spatial autoregressive process with white noise). These environmental values recurrently 'mix' at confluences considering the relative size of joining tributaries (see Methods). Therefore, our simulation resembles natural processes of how branching river networks create diverse habitats in a metacommunity. We used `r n_scenario` sets of parameter combinations with varied ecological (dispersal and competition strength) and landscape scenarios (the environmental variation at headwaters and the degree of local environmental noise). Under each scenario, we simulated metacommunity dynamics in `r max(dat_sim$n_rep)` branching networks that have different ecosystem size (the number of habitat patches: `r min(dat_sim$n_patch)` to `r max(dat_sim$n_patch)`) and complexity (branching probability: `r round(min(dat_sim$p_branch), 2)` to `r round(max(dat_sim$p_branch), 2)`) to examine their effects on species diversity.

@@

$$
\begin{align}
\mu_{z,k} &= \omega(\rho \mu_{z,k+1}^R + \epsilon_{k}^R) + (1-\omega)(\rho \mu_{z,k+1}^L + \epsilon_{k}^L) \\
\omega &= \frac{A_{k+1}^R}{A_{k+1}^R + A_{k+1}^L}
\end{align}
$$
where $R$ and $L$ denote "right" and "left" branches, respectively, and $A_k$ is the number of upstream habitat patches at longitudinal position $k$ (akin to upstream watershed area). With this expression, larger tributaties have a greater influence on the downstream environment, as observed in natural river networks[@bendaNetworkDynamicsHypothesis2004].

@@

We allowed interspecific variation in niche optimum and width, which were drawn from uniform distributions as $\mu_i \sim Unif(-1, 1)$ and $\sigma_{niche,i} \sim Unif(0.1, 1)$. We used fixed values for the following parameters: maximum reproductive number ($r_{0,i} =$ 4), niche cost ($\nu =$ `r unique(dat_sim$niche_cost)`), the degree of temporal fluctuation in abiotic environments ($\sigma_z =$ `r unique(dat_sim$sd_env)`), and the degree of spatial autocorrelation in temporal environmental dynamics ($\phi =$ `r unique(dat_sim$phi)`).

@@

We considered watersheds as independent replicates of metacommunities if the river network flows into one of the following: the ocean, a large lake ($\ge$ 10 km² in the areal area) or a large river that may represent lentic habitats ($\ge$ 5000 km² in watershed area). Metacommunities with $\ge$ 10 sampling sites were used in our analysis because small sample size inflates statistical uncertainties in estimating diversity metrics.

@@

We collected data for watershed characteristics (watershed area and branching probability), climatic variables (annual mean air temperature and cumulative precipitation), and land use (% forest). We delineated watershed polygons and river lines (defined as $\ge$ 1 km^2^) using Hydrologically Adjusted Elevations of MERIT Hydro[@yamazakiMERITHydroHighResolution2019] with ArcMap 10.7 and SAGA 6.3.0 combined with an R package 'RSAGA'[@brenningRSAGASAGAGeoprocessing2018]. Watershed area was estimated based on the delineated watershed polygons using functions in an R package 'sf'[@pebesmaSimpleFeaturesStandardized2018]. Branching probability was estimated by fitting exponential distributions to histograms of branch length (unit: kilometer, river segment between successive confluences or a confluence and the outlet/upstream terminal) as $branch~length \sim Exp(\lambda)$. A derived statistical quantity $1 − e^{−\lambda}$ corresponds to the theoretical branching probability $P_b$. For each watershed, we estimated the spatial averages of air temperature and cumulative precipitation at a 1-km resolution, which were extracted from WorldClim 1.4[@hijmansVeryHighResolution2005] via R package 'raster'. We calculated % forest for each watershed using land use data in 2015 from Copernicus Global Land Service (100-m resolution)[@marcelbuchhornCopernicusGlobalLand2020].

@@

# Acknowledgements

# Author contributions
